An integrated distance sampling approach for modelling age-structured populations

### 2023-11-30 Manuscript_Nilsen_Nater.qmd
Edits to text (.qmd) and bib-file

Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005]. Over the last few decades, a tremendous progress have been made towards this end. This progress is partly driven the development and application of new field data collection methods and apporaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref). In addition, developments of new analysis methods allow us to estimate abundance and demographic rates it situations where it was not feasible before (ref). combined, these advance put us in a much better position to estimate state variables needed for population management (ref) but also estimate indexes that are relevant for large scale policy applications, e.g. such as Essential Biodiversity Variables (ref).

@@

Distance sampling (DS) models have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. In recent years, they have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015]. These latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e. the underlying vital rates. Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021]. Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying typically age- or stage-structured population model. Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters and maximizing knowledge gained through transect surveys by augmenting them with other available data [e.g. @schmidt2020]. In this study \[...\]

Here, we will present a novel integrated population model based on distance sampling line transect survey data and demographic data from marked birds.

@@

Both juveniles and adults survive from year $t$ census to year $t+1$ census with survival probability $S_t$. As ptarmigan can reproduce already at 1-year old, all survivors then produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$. The changes in densities (numbers) of juveniles and adults in the population, $D_{juv}$ and $D_{ad}$, can thus be expressed as

@@

Note that recruitment rate $R$ is defined as juveniles/adult (not juveniles/female). We also make the simplifying assumption that there is no age- or sex-dependence of vital rates, but this assumption could be relaxed by including additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020).

@@

$$
\begin{equation}
p = exp(-\frac{y^2}{2\sigma^2})
\end{equation}
$$ where $\sigma$ is the half-normal detection parameter. We assumed $\sigma$ to vary among years (index $t$) but not transect lines or animal group size. Following @moore2011, the resulting $\sigma_t$ can be used to calculate effective strip width ($esw_{t}$) and, consequently, average detection probability per line transect with a truncation distance $W$ according to:

@@

$$
\begin{equation}
  obsN_{a,j,t} \sim Poisson(\hat{p_t}*N_{a,j,t})
\end{equation}
$$ $N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$).

The third data likelihood focuses on the counts of adults ($obsAd_{j,t}$) and juveniles ($obsJuv_{j,t}$) observed during the distance sampling surveys and links them to time-specific recruitment rate:

@@

$$
\begin{equation}
  survivors_{k,t} \sim Binomial(released_{k,t}, Sk_t)
\end{equation}
$$ Here, $Sk_t$ is the seasonal survival rate, and annual survival rate, $S_t$ is calculated as $S1_t * S2_t$.

@@

### Field data (case study)

\[Add some basic information about willow ptarmigan in Norway, plus Lierne study area\]

The two datasets used for the case study were i) data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities, and ii) data from an individual based research project based on radio collared willow ptarmigan. The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).

@@

\[One or several figures of estimates vs. true values?\]

@@

#### Population density

>>>>

Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005]. Over the last few decades, a tremendous progress have been made towards this end. This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref). In addition, developments of novel statistical methods allow researchers to estimate abundance and demographic rates it situations where it was not feasible before (ref). Combined, these advance put us in a much better position to estimate state variables needed for population management (ref) or to estimate indexes that are relevant for large scale policy applications, e.g. such as Essential Biodiversity Variables (ref).

@@

Distance sampling (DS) models have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015]. In recent years, they have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015]. These latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e. the underlying vital rates. Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021]. Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying typically age- or stage-structured population model. Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters and maximizing knowledge gained through transect surveys by augmenting them with other available data [e.g. @schmidt2020].

In this study, we will present a new IPM, where we integrate data from line transect distance sampling survey data and survival data from marked birds.

@@

Both juveniles and adults survive from year $t$ census to year $t+1$ census with survival probability $S_t$. As ptarmigan can reproduce already as 1-year old, all survivors then produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$. The changes in densities (numbers) of juveniles and adults in the population, $D_{juv}$ and $D_{ad}$, can thus be expressed as

@@

Note that recruitment rate $R$ is defined as juveniles/adult (not juveniles/female). We also make the simplifying assumption that there is no age- or sex-dependence of vital rates, but this assumption could be relaxed by including additional auxiliary data [@Israelsen2020; @Sandercock2011].

@@

$$
\begin{equation}
p = exp(-\frac{y^2}{2\sigma^2})
\end{equation}
$$ where $\sigma$ is the half-normal detection parameter. We assumed $\sigma$ to vary among years (index $t$) but not between transect lines or animal group size. Following @moore2011, the resulting $\sigma_t$ can be used to calculate effective strip width ($esw_{t}$) and, consequently, average detection probability per line transect with a truncation distance $W$ according to:

@@

$$
\begin{equation}
  obsN_{a,j,t} \sim Poisson(\hat{p_t}*N_{a,j,t})
\end{equation}
$$ $N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$, and $W$ is the truncation distance).

The third data likelihood focuses on the counts of adults ($obsAd_{j,t}$) and juveniles ($obsJuv_{j,t}$) observed during the distance sampling surveys and links them to year-specific recruitment rate:

@@

$$
\begin{equation}
  survivors_{k,t} \sim Binomial(released_{k,t}, Sk_t)
\end{equation}
$$ Here, $Sk_t$ is the seasonal survival probability, and annual survival probability, $S_t$ is calculated as $S1_t * S2_t$.

@@

### Case study

The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems. In Norway, there has been a long-term decline in the willow ptarmigan abundance spanning more than a century (ref), but the last few decades the trend in abundance has fluctuated both in time and space. In Scandinavia, willow ptarmigan is a valued game species (ref), and there has been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011]. A key insight from across several study areas is the the annual recruitment rate (i.e. $R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents as alternative prey for shared predators (i.e. the Alternative Prey Hypothesis ref). Adult survival show less inter-annual fluctuations [@Israelsen2020], although spatial (and potentially temporal) variation due to e.g. harvest management is evident when comparing across studies [@Israelsen2020].

Our case study was based on an ongoing long-term research project on willow ptarmigan in Lierne municipality in Central Norway (approximately 62.4 degrees north and 13.2 degrees east). The study area is located in a sub-alpine ecosystem, and the landscape is a mosaic of open heath and shrub vegetation (dominated by Ericacea, willow shrub *Salix spp*. and dwarf birch *Betula nana*), interspersed with bogs and forest patches (mainly birch *Betula spp*.). The climate is strongly seasonal, with snow typically covering the ground from October/November through April/May.

From this study system, two datasets were used for the case study:

1.  Data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities

2.  Data from an individual based research project based on radio collared willow ptarmigan. The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).

@@

In general, the model was able to estimate both parameters of the detection model (Figure X) of the simulated data with little bias.

![](Figures/SimCheck_Detects_Temps.png)

**Figure X** Model performance on simulated data. Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability

![](Figures/SimCheck_VRs_Temps.png)

**Figure X** Vital rates estimated based on simulated data

@@

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

library(tidyverse)
library(tidybayes)

## Posterior samples added to folder "Posterior_samples". 
## .rds are included in gitignore

test <- readRDS("Posterior_samples/rypeIDSM_dHN_multiArea_realData_Lierne_tidy.rds")

test1 <- test %>% spread_draws(Mu.R[])
mean_MuR <- mean(test1$Mu.R)
ci_MuR <- quantile(test1$Mu.R, probs = c(0.025, 0.975))

```

Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set. As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021. The increase was most distinct from 2016 and onward. Across years, mean survival probability for August - January ($S_1$) was estimated at X (95% C.I = ), survival probability for February - July ($S_1$) was estimated as X (95% C.I = x), and annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at X. Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021. Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability. While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(mean_MuR, 1)` (95% C.I. = `r as.numeric(round(ci_MuR[1], 1))` - `r as.numeric(round(ci_MuR[2], 1))` ).
